---
title: "Final Project"
author: "Shashank"
date: "06/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(dplyr)
library(ggplot2)
```

### Introduction
The goal of the project is to analyze factors affecting life expectancy. Majorly focusing on the relationship of health care (Expenditure, Immunization) and education with life expectancy

```{r}
life=read.csv("Life Expectancy cleaned.csv",sep=",",header = TRUE)
```

# cleaning 
```{r,message=FALSE}
#life = read_csv("../Final Project/Life Expectancy Data.csv")

# names(life)[4] = "Life_expectancy" 
# names(life)[8] = "percentage_expenditure" 
# names(life)[9] = "HepatitisB"
# names(life)[14] ="Total_expenditure"
# # names(life)
# selected_names = c("Country", "Year", "Status", "Life_expectancy", "percentage_expenditure", "Polio", "Diphtheria", "Total_expenditure", "Schooling")
# life = life%>% dplyr::select(selected_names)
# 
# life=na.omit(life)
```


```{r}
#install.packages("countrycode")
library(countrycode)
#life1=life
#life1 = data.frame(life)
life=data.frame(life)
life$continent <- countrycode(sourcevar = life[, "Country"],
                            origin = "country.name",
                            destination = "continent")
sa = c("Argentina", "Brazil", "Chile","Colombia","Ecuador", "Guyana", "Paraguay", "Peru", "Suriname", "Uruguay","Bolivia (Plurinational State of)","Venezuela (Bolivarian Republic of)")
na=c("Belize","Canada","Costa Rica","Dominican Republic","El Salvador", "Guatemala", "Haiti", "Honduras", "Jamaica", "Mexico","Nicaragua","Panama","Trinidad and Tobago","Bahamas","Barbados","Saint Lucia","Antigua and Barbuda","Cuba","Grenada","Saint Vincent and the Grenadines")
#life2=data.frame(life1)
life[which(life$Country %in% sa),]$continent="South America"
life[which(life$Country %in% na),]$continent="North America"

```

## Trend for Life expectancy 

#### As the year increases the life expenctancy is increasing. Developed countries have a higher life expectancy

```{r, message= False}
life1=life[,c("Year","continent","Life_expectancy")]
life_mean=aggregate(Life_expectancy~Year+continent,data=life1,mean)

ggplot(life_mean %>% filter(Year < 2014) ,aes(x = Year , y = Life_expectancy))+
  geom_point(color=" gray")+
  geom_smooth(method = 'glm',color="dark blue", se=FALSE)+ 
  labs(title = "Trend of Life expenctancy over the years by Continents",
           #caption = "HPI: House Price Index",
    x=" Year",y="Life Expectancy")+
    theme_bw() +facet_wrap(~continent)

```

### Health Care related factors:
1. Expenditure in health care
2. Polio


### Relation of life expectancy with health care factors

```{r}
ggplot(life %>% filter(Year < 2014) ,aes(x = Total_expenditure, y = Life_expectancy))+
  geom_point(color=" gray") + #scale_x_log10()+
  geom_smooth(method="glm",color="dark blue", se=FALSE)+ 
  #xlim(c(0,50))+
  labs(title = "Life Expectancy vs Percentage of Expenditure in health care",
           #caption = "HPI: House Price Index",
    x=" Percentage_expenditure",y="Life Expectancy")+
    theme_bw() +facet_wrap(~continent)

```

#### For one year as Brad suggested, kind of have similar trend with all the data points

```{r}
ggplot(life %>% filter(Year == 2014) ,aes(x = Total_expenditure, y = Life_expectancy))+
  geom_point(color=" gray") + #scale_x_log10()+
  geom_smooth(method="glm",color="dark blue", se=FALSE)+ 
  #xlim(c(0,50))+
  labs(title = "Life Expectancy vs Percentage of Expenditure in health care",
           #caption = "HPI: House Price Index",
    x=" Percentage_expenditure",y="Life Expectancy")+
    theme_bw() +facet_wrap(~continent)

```
### Wanted to check the relationship of percentage expenditure solely based on Africa 

```{r}
ggplot(life %>% filter(continent == 'Africa') ,aes(x = Total_expenditure, y = Life_expectancy))+
  geom_point(color=" gray") + #scale_x_log10()+
  geom_smooth(method="glm",color="dark blue", se=FALSE)+ 
  #xlim(c(0,50))+
  labs(title = "Life Expectancy vs Percentage of Expenditure in health care",
           #caption = "HPI: House Price Index",
    x=" Percentage_expenditure",y="Life Expectancy")+
    theme_bw() + facet_wrap(~cut(Year,3))

```

```{r}
ggplot(life %>% filter(continent == 'Africa',  Total_expenditure > 5,Life_expectancy<50) ,aes(x = Total_expenditure, y = Life_expectancy))+
  geom_point(color=" gray") + #scale_x_log10()+
  geom_smooth(method="glm",color="dark blue", se=FALSE)+ 
  #xlim(c(0,50))+
  labs(title = "Life Expectancy vs Percentage of Expenditure in health care",
           #caption = "HPI: House Price Index",
    x=" Percentage_expenditure",y="Life Expectancy")+
    theme_bw() + facet_wrap(~cut(Year,3))
```

```{r}
highexp_lowexp05 = life %>% filter(continent == 'Africa',  Total_expenditure > 5,Life_expectancy<50, Year<2005)
highexp_lowexp05_Country = unique(highexp_lowexp05$Country)
```
```{r}
highexp_lowexp05_Country
```

```{r}
life %>% filter(Country %in% highexp_lowexp05_Country) %>% ggplot(aes(y=Life_expectancy, x=Total_expenditure, color=Year))+
                                                                      # , color=Country, group = Country))+
  geom_point()+
  facet_wrap(~ Country)+
  theme_bw()
```

```{r}
africa_high_exp = life %>% filter(continent == 'Africa', Total_expenditure > 10)
africa_high_exp = africa_high_exp[c("Country", "Year", "Life_expectancy", "Total_expenditure", "Schooling", "Polio")]
africa_high_exp %>% filter(Year<2006)
```

```{r}
africa_high_exp %>% group_by( Country)  %>% summarise(meanLE=mean(Life_expectancy),medLE = median(Life_expectancy)  ,n=n())
```

```{r}
africa_high_exp %>% group_by(Country)  %>% summarise(meanSch=mean(Schooling),medSch = median(Schooling), minSch=min(Schooling),maxSch = max(Schooling) )

```


# % change of life expectancy and expenditure in health care


```{r}
cpi_00 <- subset(life, life$Year == 2000)#[c('NAME','value')]
cpi_14 <- subset(life, life$Year == 2014)#[c('NAME','value')]

cpi_combined=data.frame(cpi_00)
colnames(cpi_combined)[colnames(cpi_combined) == "Life_expectancy"] <- "Life_expectancy_2000"
colnames(cpi_combined)[colnames(cpi_combined) == "Total_expenditure"] <- "Total_expenditure_2000"
colnames(cpi_combined)[colnames(cpi_combined) == "Status"] <- "Status_2000"

cpi_combined <- merge(cpi_combined, cpi_14, by = c("Country"))
colnames(cpi_combined)[colnames(cpi_combined) == "Life_expectancy"] <- "Life_expectancy_2014" 
colnames(cpi_combined)[colnames(cpi_combined) == "Total_expenditure"] <- "Total_expenditure_2014"
colnames(cpi_combined)[colnames(cpi_combined) == "Status"] <- "Status_2014"

#2005
cpi_combined$cpi_life_2005<- round((cpi_combined$Life_expectancy_2014 - cpi_combined$Life_expectancy_2000)*100/cpi_combined$Life_expectancy_2000)

cpi_combined$cpi_expenditure_2005<- round((cpi_combined$Total_expenditure_2014 - cpi_combined$Total_expenditure_2000)*100/cpi_combined$Total_expenditure_2000)
```



# years 2000-2014
```{r}
ggplot( aes(x = cpi_expenditure_2005, y =cpi_life_2005),data=cpi_combined)+xlim(c(0,100))+geom_point(color=" gray") +geom_smooth(method='glm',se=FALSE)+facet_wrap(~continent.x,labeller = label_both) +theme_bw() + labs(title = "Snapshot %change in expenditure vs %change in expectancy  from 2000-2014",
    x="Change in Percentage_expenditure",y="Change in Life Expectancy")
```


#### Polio

```{r}
ggplot( data=life %>% filter(Year < 2014),aes(x = Polio , y =Life_expectancy))+geom_point(color=" gray")+ggtitle('Polio vs Life expectancy faceted on continents')+geom_smooth(method='lm',se=FALSE)+facet_wrap(~continent,labeller = label_both) + theme_bw() 
```


```{r}
life%>%filter(Polio<25, Status=="Developed")
```
```{r}
ggplot(life, aes(x=Total_expenditure))+geom_histogram()
```

```{r}
life%>%filter(Total_expenditure<1, Status=="Developed")

```

```{r}
# life%>%filter(Total_expenditure<1, Status=="Developing")
```


```{r}
ggplot(life, aes(x=Schooling))+geom_histogram()

```


### Schooling

```{r}
ggplot( aes(x = Schooling, y =Life_expectancy),data=life)+
    geom_point(color=" gray")+
    facet_wrap(~continent,labeller = label_both)+
  geom_smooth(method='glm',color="dark blue", se=FALSE)+
  labs(title = "Life Expectancy vs No. of year of education",
           #caption = "HPI: House Price Index",
    x=" Years of schooling",y="Life Expectancy")+
    theme_bw() +facet_wrap(~continent)
```


### Correlation plot

```{r}
library(GGally)
ggpairs(life[c("Year", "Polio", "Diphtheria", "Total_expenditure", "Schooling" )])
```


```{r}
model.lm = lm(Life_expectancy ~Status  + Polio +Diphtheria+ Total_expenditure+Schooling+continent + continent:Total_expenditure + continent:Schooling, data=life)
summary(model.lm)
```

```{r}
AIC(lm(Life_expectancy ~ Status  + Polio +Diphtheria+ Total_expenditure+Schooling+continent + continent:Total_expenditure + continent:Schooling, data=life))
```


```{r}
model.lm = lm(Life_expectancy ~Status  + Polio +Diphtheria+ Total_expenditure+Schooling+continent, data=life)
summary(model.lm)
```

```{r}
AIC(model.lm)
```

```{r}
model.lm = lm(Life_expectancy ~ Polio +Diphtheria+ Total_expenditure+Schooling+continent + continent:Total_expenditure + continent:Schooling, data=life)
# summary((model.lm))
library(broom)
model.lm.aug = augment(model.lm)
ggplot(model.lm.aug, 
       aes(x=Total_expenditure, y=.resid)
       )+
  geom_point(shape=16, alpha=0.8)+
  geom_smooth(method = "lm")+
  facet_wrap(~continent)+
  theme_bw()+
  ggtitle("Residuals against Expenditure")
```

```{r}

model.lm = lm(Life_expectancy ~ Polio + Total_expenditure + Schooling + 
                continent + continent:Total_expenditure + continent:Schooling, data=life)

# summary((model.lm))
library(broom)
model.lm.aug = augment(model.lm)
ggplot(model.lm.aug, 
       aes(x=Schooling, y=.resid)
       )+
  geom_point(shape=15, alpha=0.8)+
  geom_smooth(method = "lm")+
  facet_wrap(~continent)+
  theme_bw()+
  ggtitle("Residuals against Polio")
```


```{r}
summary((model.lm))
AIC(model.lm)
```

```{r}
model.lm = lm(Life_expectancy ~ Polio +Diphtheria+ Total_expenditure+Schooling+continent + continent:Total_expenditure + continent:Schooling, data=life)
# summary((model.lm))
library(broom)
model.lm.aug = augment(model.lm)
ggplot(model.lm.aug, 
       aes(x=Polio, y=.resid)
       )+
  geom_point(shape=15, alpha=0.8)+
  geom_smooth(method = "lm")+
  facet_wrap(~continent)+
  theme_bw()+
  ggtitle("Residuals against Polio")
```


```{r}
life.grid=expand.grid(Polio=seq(min(life$Polio),max(life$Polio),1),Total_expenditure=seq(min(life$Total_expenditure),max(life$Total_expenditure),0.25),Schooling=seq(min(life$Schooling),max(life$Schooling),1),continent=c("Asia","Europe","Africa","North America","South America","Oceania"),Status=c("Developed","Developing"))

```

```{r}
life.pred = predict(model.lm,
type = "response", newdata = life.grid)
life.pred.df = data.frame(life.grid,
life.prob = as.vector(life.pred))
```

```{r}
life.pred.df
```

```{r}
# ggplot(life.pred.df, aes(x=Polio, y=life.prob, color=Status))+
#   geom_point()+
#   facet_wrap(~continent)
  
```

