---
title: "EDA_Project"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(dplyr)
library(ggplot2)
```

### Introduction
The goal of the project is to analyze factors affecting life expectancy. Majorly focusing on the relationship of health care (Expenditure, Immunization) and education with life expectancy

```{r}
life=read.csv("Life Expectancy Data.csv",sep=",",header = TRUE)
```


# cleaning 
```{r,message=FALSE}
#life = read_csv("../Final Project/Life Expectancy Data.csv")

names(life)[4] = "Life_expectancy" 
names(life)[8] = "percentage_expenditure" 
names(life)[9] = "HepatitisB"
names(life)[14] ="Total_expenditure"
# names(life)
selected_names = c("Country", "Year", "Status", "Life_expectancy", "percentage_expenditure", "Polio", "Diphtheria", "Total_expenditure", "BMI", "GDP", "Population", "Schooling", "HepatitisB")
life = life%>% dplyr::select(selected_names)

life=na.omit(life)
```

## Trend for Life expectancy 

#### As the year increases the life expenctancy is increasing. Developed countries have a higher life expectancy

```{r, message= False}
life1=life[,c("Year","Status","Life_expectancy")]
life_mean=aggregate(Life_expectancy~Year+Status,data=life1,mean)

ggplot(life_mean,aes(x = Year , y = Life_expectancy))+
  geom_point(color=" gray")+
  geom_smooth(color="dark blue", se=FALSE)+ 
  labs(title = "Trend of Life expenctancy over the years",
           #caption = "HPI: House Price Index",
    x=" Year",y="Life Expectancy")+
    theme_bw() +facet_wrap(~Status)

```

### Relation of life expectancy with health care factors


```{r}
ggplot(life,aes(x = Total_expenditure , y = Life_expectancy))+
  geom_point(color=" gray") + #scale_x_log10()+
  geom_smooth(method="glm",color="dark blue", se=FALSE)+ 
  #xlim(c(0,50))+
  labs(title = "Life Expectancy vs Percentage of Expenditure in health care",
           #caption = "HPI: House Price Index",
    x=" Percentage_expenditure",y="Life Expectancy")+
    theme_bw() +facet_wrap(~Status)

```

# % change of life expectancy and expenditure in health care


```{r}
cpi_00 <- subset(life, life$Year == 2000)#[c('NAME','value')]
cpi_14 <- subset(life, life$Year == 2013)#[c('NAME','value')]

cpi_combined=data.frame(cpi_00)
colnames(cpi_combined)[colnames(cpi_combined) == "Life_expectancy"] <- "Life_expectancy_2000"
colnames(cpi_combined)[colnames(cpi_combined) == "Total_expenditure"] <- "Total_expenditure_2000"
colnames(cpi_combined)[colnames(cpi_combined) == "Status"] <- "Status_2000"

cpi_combined <- merge(cpi_combined, cpi_14, by = c("Country"))
colnames(cpi_combined)[colnames(cpi_combined) == "Life_expectancy"] <- "Life_expectancy_2014" 
colnames(cpi_combined)[colnames(cpi_combined) == "Total_expenditure"] <- "Total_expenditure_2014"
colnames(cpi_combined)[colnames(cpi_combined) == "Status"] <- "Status_2014"

#2005
cpi_combined$cpi_life_2005<- round((cpi_combined$Life_expectancy_2014 - cpi_combined$Life_expectancy_2000)*100/cpi_combined$Life_expectancy_2000)

cpi_combined$cpi_expenditure_2005<- round((cpi_combined$Total_expenditure_2014 - cpi_combined$Total_expenditure_2000)*100/cpi_combined$Total_expenditure_2000)
```



# years 2000-2014
```{r}
ggplot( aes(x = cpi_expenditure_2005, y =cpi_life_2005),data=cpi_combined)+xlim(c(0,100))+geom_point(color=" gray") +geom_smooth(method='glm',se=FALSE)+facet_wrap(~Status_2000,labeller = label_both)+ggtitle('Snapshot %change in expenditure vs %change in expectancy  from 2000-2014')+theme_bw() 
```

### Relationship of life expectancy with Immunization variable

#### Life expectancy vs Polio Immunization

```{r}
ggplot( aes(x = Polio, y =Life_expectancy),data=life)+geom_point(color=" gray")+ggtitle('Polio vs Life_expectancy faceted on status')+geom_smooth(method='lm',se=FALSE)+facet_wrap(~Status,labeller = label_both) + theme_bw() 
```

```{r}
ggplot(life, aes(y=Total_expenditure, x=Year, color=Status, group=Status))+
  geom_jitter(width=0.15, height = 0, alpha=0.9)+
  facet_wrap(~cut(Polio, 5), ncol=2, labeller = label_both)+
  
    ggtitle("Expenditure on healthcare & Polio immunization over the years", subtitle = "Expenditure on healthcare as a %GDP per capita")+
  theme_bw()

  # scale_y_log10()
  # geom_smooth()
```

#### Life expectancy vs HepatitisB Immunization

```{r}
ggplot( aes(x = HepatitisB, y =Life_expectancy),data=life)+geom_point(color=" gray")+ggtitle('HepatitisB vs Life_expectancy faceted on status')+geom_smooth(method='lm',se=FALSE)+facet_wrap(~Status,labeller = label_both) + theme_bw() 
```

#### Life expectancy vs Education

```{r}
ggplot( aes(x = Schooling, y =Life_expectancy),data=life)+
    geom_point(color=" gray")+
    facet_wrap(~Status,labeller = label_both)+
  geom_smooth(method='glm',color="dark blue", se=FALSE)+
  labs(title = "Life Expectancy vs No. of year of education",
           #caption = "HPI: House Price Index",
    x=" Years of schooling",y="Life Expectancy")+
    theme_bw() +facet_wrap(~Status)
```


```{r}
ggplot(life, aes(x=Schooling, y=Life_expectancy, color=Status, group=Status))+
  geom_point()+
  geom_smooth(method="lm", col="black", se=F)+
  facet_wrap(~cut_number(Year, 4))+
  ggtitle("Life expectancy over the years for Developing and Developed countries")+
  theme_bw()
  # scale_y_l
```

Positive corr between scholing and life expectancy. 
Strength of corr to be calculated...
Developed countries all have higher rates of education and higher life expectancy [consistent over the years]









